---
title: "Did the Canadian wildfires have an effect on bird migration?"
author: "Claire"
date: "2023-09-18"
output: html_document
---

```{r libraries, include=FALSE, echo=FALSE}
library(dplyr)
library(tidyverse)
library(broom)
library(ggplot2)
library(viridis)
library(hrbrthemes)
library(plotly)
```



```{r get data, echo=FALSE}
wildfires0 <- read.table(paste0(dirname(getwd()), '/data/NFDB_point_txt/NFDB_point_20220901.txt'), sep=",",fill=TRUE,quote = "", header=TRUE)
bird0 <- read.csv(paste0(dirname(getwd()), '/data/Complete_data.csv'))
```


## Baseline model

We set up a basic linear model to evaluate the impact of wildfires on species abundance among migrating species. 
We match up the years and states that we need on both datasets. And only collect years in which we see a wildfire occurring at breeding season (May-June-July). Since the wildfire data covers Canadian states, we'll keep those only. 
```{r match data, echo=FALSE}
key_table <- read.table(paste0(dirname(getwd()), '/data/NFDB_point_txt/id_key.txt'), header=TRUE, sep = ',')
# Get breeding season
wildfires <- wildfires0[wildfires0$MONTH %in% c(5:7),]
# Match up the years
wildfires <- wildfires[wildfires$YEAR %in% unique(bird0$Year),]
wildfires <- wildfires[wildfires$SRC_AGY2 %in% unique(key_table$ID_CAN),]

bird <- bird0[bird0$Year %in% unique(wildfires$YEAR),]
# Match up states 
bird <- bird[bird$StateNum %in% key$ID_BBS,]
```


First model: aggregated across all species, $\mathtt{bird\_pop}_{\mathtt{year, state}} \sim \beta_0 + \beta_1 \mathtt{wildfire}_{\mathtt{year, state}}$. 



```{r bird pop plot, echo=FALSE}
bird_pop <- bird %>%
  group_by(Year, StateNum) %>%
    summarise(tot = sum(SpeciesTotal)) %>%
      full_join(key_table,by=c("StateNum"="ID_BBS"))
ggplot(bird_pop, aes(x=Year, y=tot, group=factor(StateNum), color=factor(name))) +
    geom_line() + scale_color_viridis(discrete = TRUE) +ggtitle('All species across Canadian states from 1966 to 2021')+theme_ipsum() + xlab("Year") +ylab("Species count (total)")
```


```{r wildfires plot, echo=FALSE}
wildfires$SIZE_HA <- as.numeric(wildfires$SIZE_HA)
fires_per_year <- wildfires %>% 
  group_by(YEAR, SRC_AGY2) %>% 
    summarise(inc = sum(SIZE_HA)) %>%
      full_join(key_table,by=c("SRC_AGY2"="ID_CAN"))

ggplot(fires_per_year, aes(x=YEAR, y=inc, group=factor(SRC_AGY2), color=factor(name))) +
    geom_point() + scale_color_viridis(discrete = TRUE) +ggtitle('All incidence of fires across Canadian states from 1966 to 2021')+xlab('Year')+ylab('Fire incidence')+theme_ipsum()
```
```{r}
combined_table <- bird_pop %>% full_join(fires_per_year, by= c("Year"="YEAR", "ID_CAN"="SRC_AGY2"))
combined_table$inc[is.na(combined_table$inc)] <- 0
```


```{r model}
linear_model <- lm(data=combined_table, tot ~ inc +factor(ID_CAN)+ Year)
summary(linear_model)
```
```{r}
plot(linear_model)
save.image('/Users/clairehe/Desktop/phd year 2/ADS_TA_template/proj1/week3/img.png') 
```
Non linearity problem. Trend has almost a quadratic curve. 

```{r}
linear_model2 <- lm(data=combined_table, tot ~ inc*factor(ID_CAN)+Year)
summary(linear_model2)
anova(linear_model2, linear_model)
```


```{r}
plot_ly(x=combined_table$inc, y=combined_table$tot, z=combined_table$Year, type="scatter3d", mode="markers", color=combined_table$name.y) %>%  layout(scene = list(xaxis = list(title = 'fire incidence'),
                     yaxis = list(title = 'Species total count'),
                     zaxis = list(title = 'Year')))
```

```{r}
ggplot(combined_table[combined_table$Year == '2013',], aes(x=inc, y=tot, color=factor(name.y))) +geom_point()+xlab('Fire incidence')+ylab('Total species count for year 2013')+theme_ipsum()
ggplot(combined_table, aes(x=inc, y=tot, color=factor(name.y))) +geom_point()+xlab('Fire incidence')+ylab('Total species count')+theme_ipsum()
ggplot(combined_table[combined_table$ID_CAN=='NL',], aes(x=inc, y=tot)) +geom_point()+xlab('Fire incidence')+ylab('Total species count in Newfoundland and Labrador')+theme_ipsum() 
```
